Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-14 manuscript/climate_sensitivity_SI.Rmd
adjusted top/best model terminology
I've implemented your suggestion, @mcgregorian1 . This closes #103
figure/table numbering
#104

\newpage
```{r table S7, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
kable(multivar_table, booktabs = TRUE, caption = "Table S7. Summary of top full models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
                     " "," ")) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 8, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 20, 30, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(8, bold = T) %>%
  row_spec(20, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold).

@@

\newpage
![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)

\newpage
![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

>>>>

\newpage
```{r table S7, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
kable(multivar_table, booktabs = TRUE, caption = "Table S7. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
                     " "," ")) %>%
  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 8, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 20, 30, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(5, bold = T) %>%
  row_spec(8, bold = T) %>%
  row_spec(20, bold = T) %>%
  kable_styling()
```
Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold).

@@

\newpage
![Figure S4: **PLACEHOLDER** ](tables_figures/publication/independent_variables_by_species.png)

\newpage
![Figure S5: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)

\newpage
![Figure S6: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
moved subsection on identifying drought years up in methods
figure/table numbering
#104

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA. First, we focus on the role of tree height and its interaction with microenvironment. 
We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

@@

Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park. Elevations range from 273 to 338 m above sea level with a topographic relief of 65m [@bourg_initial_2013]. Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014). Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig. **S1**). Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

For each cored tree, we combined tree-ring records and allometric equations of bark thickness to reconstruct DBH for the years 1950-2009. 
Prior DBH was estimated using the following equation:

@@

Hydraulic traits were collected in August 2018 (Tables 2-3). We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info. We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al. (2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **2**) and associated traits at a single site [see also @elliott_forest_2015]. 
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. 2). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.  
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S#**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
A second reason is microclimate...

>>>>

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2). 
First, we focus on the role of tree height and its interaction with microenvironment. 
We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

@@

Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W; Fig. **S1**) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park. Elevations range from 273 to 338 m above sea level with a topographic relief of 65m [@bourg_initial_2013]. Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014). Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 3).

@@

We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig. **S1**). Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
The resulting chronologies (Fig. **1a**) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

For each cored tree, we combined tree-ring records and allometric equations of bark thickness to reconstruct DBH for the years 1950-2009. 
Prior $DBH$ was estimated using the following equation:

@@

Hydraulic traits were collected in August 2018 (Tables 2-3; Fig. **S4**). We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot. Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken. Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed. Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea]. $LMA$ was calculated as the ratio of leaf dry mass to fresh area. $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves. Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement. We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$). Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA). Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].

@@

For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info. We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al. (2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **2**) and associated traits at a single site [see also @elliott_forest_2015]. 
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. 2). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.  
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
A second reason is microclimate...
